---
title: "Data Visualization Report"
author: "Musaib Syed"
date: "`r Sys.Date()`"
output: html_document
---

## **Introduction**

This report explores a dataset using **ggplot2** for data visualization. The dataset used in this example is `mtcars`, a built-in dataset in R that contains information about different car models.

---

## **Load Required Packages**

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
```

---

## **Load and Explore the Dataset**

```{r}
data(mtcars) # Load dataset
head(mtcars) # Display first few rows
```

### **Summary Statistics**

```{r}
summary(mtcars)
```

The dataset contains multiple numerical variables describing various car features, such as **miles per gallon (mpg)**, **horsepower (hp)**, **weight (wt)**, and **number of cylinders (cyl)**.

---

## **Data Visualization: MPG vs Horsepower with Cylinder Categories**

### **Objective:**
- **Compare miles per gallon (MPG) against horsepower (HP).**
- **Color points by number of cylinders (`cyl`).**
- **Size of points represents weight (`wt`).**
- **Trend line added for better readability.**

```{r}

ggplot(mtcars, aes(x = hp, y = mpg, color = factor(cyl))) +
  geom_point(aes(size = wt), alpha = 0.7) +  # Apply size only to points
  geom_smooth(aes(group = factor(cyl)), method = "lm", se = FALSE, linetype = "dashed", color = "black", linewidth = 1) +  # Fix group issue
  labs(
    title = "Miles Per Gallon vs Horsepower",
    subtitle = "Data grouped by Number of Cylinders",
    x = "Horsepower (hp)",
    y = "Miles Per Gallon (mpg)",
    color = "Cylinders",
    size = "Weight (wt)"
  ) +
  theme_minimal() +  # Clean theme
  scale_color_manual(values = c("red", "blue", "green"))  # Custom color scheme


```

---

## **Plot Explanation**

### **What the Plot Shows:**
- **MPG decreases as horsepower increases.**
- **Cars with fewer cylinders (4-cyl) tend to have higher MPG.**
- **Heavier cars (larger points) are generally less fuel-efficient.**
- **A trend line (`geom_smooth()`) confirms a negative correlation between MPG and HP.**

### **Visualization Best Practices Applied:**
✔ **Color coding** (`color=factor(cyl)`) to differentiate cylinder categories.  
✔ **Size encoding** (`size=wt`) to represent weight visually.  
✔ **Added a trend line** (`geom_smooth(method="lm")`) for clarity.  
✔ **Clear labels and title** (`labs()`).  
✔ **Minimalist theme** (`theme_minimal()`) to reduce clutter.  
✔ **Custom colors** (`scale_color_manual()`) to enhance readability.  

---

## **Conclusion**

This visualization effectively highlights how **MPG decreases with higher horsepower** and how **cylinder count and weight** affect fuel efficiency.

To **render** this file, save it as `.Rmd` and use **Knit** in RStudio to generate an HTML report.


